%Acknowledgements are the author's statement of gratitude to and
%recognition of the people and institutions who helped the author's
%research and writing.

\begin{center}
\vspace*{52pt}
{\Large \textbf{ACKNOWLEDGEMENTS}}
\end{center}

I thank the Valdar lab for being a supportive and enriching venue in which to work and learn these past five years.
I appreciate the efforts Will has made to create this lab environment by recruiting talented students such as Greg Keele, Paul Maurizio, Dan Oreper, Wes Crouse, Yanwei Cai, and Kathie Sun.

% I have been allowed the time to think about what I'm doing in a way that is not guaranteed in graduate school.

I thank the BCB program.
They built it and we came.
Tim Elston, Jonathan Cornett, and Cara Marlow were supportive in a thousand little ways that helped me stay focused on my academics.

I thank the MD-PhD program.
The ``big picture'' guidance and support I've received from Gene Orringer, Toni Darville, Mohanish Deshmukh, Alison Regan, Carol Herion has kept me moving forward in a productive direction no matter how thick the morass of graduate school felt.

I thank my family.
My mom, dad, and brother have been patient with me when I needed it and gave me a little kick sometimes when I needed that too.
% Conversations with my mother begin with the axiom that her sons are perfect in every way.
% By some miracle, they end on the 
% Conversations with my father tend the other direction with a clear-headed analysis of where I've gone wrong --- a hard truth.

I especially want to thank my grandparents.
From childhood, my mom's parents shared with me a love of reading and writing.
Every step of the way, the importance of those skills seems to compound.
% Most of the time my grandmother came over to visit, she brought some books from the bookstore where she worked.
% I should have read them!
My dad's parents brought in another piece of the puzzle.
With visits to the Franklin institute and a home experimenter kit, they kindled my enthusiasm for science.
I'm thrilled to be where I am and I am deeply grateful to them for helping me get here.

\clearpage
